---
title: 'Assignment 1: Activity Monitoring'
author: "IB"
date: "June 1, 2015"
output: html_document
---
```{r, warning = FALSE}
require (plyr)
require (ggplot2)
```

#I. Loading and preprocessing the data.
```{r}
data <- read.csv("activity.csv", header = TRUE)
head(data)
summary(data)
```

#II. What is mean total number of steps taken per day?

###1. Calculate the total number of steps taken per day.
```{r}
steps.day <- ddply(data, "date", summarise, steps = sum(steps))
head(steps.day)
```

###2. Make a histogram of the total number of steps taken each day.
```{r, echo=TRUE}
ggplot(steps.day, aes(x=steps)) + geom_histogram(colour = "black", 
                                                 fill = "skyblue", 
                                                 binwidth = 1000) + 
        xlab("Total Steps per Day") +
        ylab("Frequency")+
        theme_classic()
```

###3. Calculate and report the mean and median of the total number of steps taken per day.
```{r}
mean.steps <- round(mean(steps.day$steps,na.rm=TRUE ), digits=0)
mean.steps

median.steps <- round(median(steps.day$steps, na.rm=TRUE), digits=0)
median.steps
```
####The mean of the total number of steps taken per day is `r paste(mean.steps)`. 
####The median of the total number of steps taken per day is `r paste(median.steps)`.

#III. What is the average daily activity pattern?

###1.Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
interval.day <- ddply(data, "interval", summarise, steps = mean(na.omit(steps)))
head(interval.day)
```

```{r, echo=TRUE}
ggplot(interval.day, aes(x = interval, y = steps)) + 
    geom_line(size=1, colour = "red") + theme_bw()+
        xlab("Interval") +
        ylab("Average Number of Steps")
```

###2.Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
step.max = round(interval.day[(which.max(interval.day$steps)),], digits=0)
step.max
```

#####The `r step.max$interval`th interval contains the average maximum number of `r step.max$steps` steps. 
####Same results reported as a table:

Interval | Steps
-------- | -----
835      | 206  

 Table: __Interval with Average Maximum Steps__ 

#III. Imputing missing values.
 
###1.Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
total.NA <- sum(is.na(data))
total.NA
``` 
####The total number of rows with NAs in the dataset is `r total.NA`.

###2.Devise a strategy for filling in all of the missing values in the dataset.Use the mean for that 5-minute interval.
```{r}
newdata <- merge(data, interval.day, by = "interval")

for (i in 1:nrow(newdata)){
 if (is.na(newdata$steps.x[i])) {
   newdata$steps.x[i] <- newdata$steps.y[i]
 }}
``` 

###3.Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
newdata$steps.y=NULL
names(newdata)[2]<-"steps"
newdata <- newdata[order(newdata$date, newdata$interval),] 
newdata <- newdata[c(2,3,1)]
head(newdata)
``` 

###4.Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 

```{r, echo=TRUE}
newsteps.day <- ddply(newdata, "date", summarise, steps = sum(steps))

ggplot(newsteps.day, aes(x=steps)) + geom_histogram(colour = "black", 
                                                 fill = "cornsilk", 
                                                 binwidth = 1000) + 
        xlab("Total Steps per Day") +
        ylab("Frequency")+
        theme_classic()
```

```{r}
mean.steps2 <- round(mean(newsteps.day$steps,na.rm=TRUE ), digits=0)
mean.steps2
median.steps2 <- round(median(newsteps.day$steps, na.rm=TRUE), digits=0)
median.steps2
```
####The mean of the total number of steps taken per day is `r paste(mean.steps2)`. 
####The median of the total number of steps taken per day is `r paste(median.steps2)`.

###Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

The mean is the same as in the first part: 10,766. The median increased from 10,765 to 10,766.

The new histogram is very similar to the histogram from the first part. However, the highest frequency increased and moved to the right from 10,000 to 11,000 steps per day.

#IV. Are there differences in activity patterns between weekdays and weekends?

###1.Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
newdata$day <- ifelse(weekdays(as.Date(newdata$date)) %in% c("Saturday", "Sunday"), "weekend", "weekday")
head(newdata)
```

###2.Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r, echo=TRUE}
daydata <- ddply(newdata, .(interval, day), summarise, steps = mean(steps))
head(daydata)

tsplot <- ggplot(daydata, aes(x = interval, y = steps, colour = day)) + 
    geom_line(size=1) + facet_grid(day ~ .) + theme_bw() 

tsplot
``` 

